---
title: "Findings and data bulletproofing"
author: "Dillon Bergin"
date: "5/28/2025"
execute:
  output: false
  echo: false
format:
  html:
    code-tools: true
    theme: cosmo
    toc: true
---
```{r}
library(tidyverse)
library(here)
library(janitor)
library(ggplot2)
library(plotly)
library(DT)
library(lubridate)
```


```{r}
present <- read.csv(here("data", "processed","analysis", "present.csv")) %>% 
  select(-X, -...1)
prior <- read.csv(here("data", "processed","analysis", "prior.csv")) %>% 
  select(-X, -...1)
reinstated <- read.csv(here("data", "processed","analysis", "reinstated.csv")) %>% 
  select(-X, -...1)
certified_directory <- read.csv(here("data", "processed","analysis", "certified_directory.csv")) %>% 
  select(-X)
```

# How have decertifications changed over time?
## Decertifications increased after law change
### Sentences in the article 

> 1: "After decades when few law enforcement officers in Virginia lost the right to wear a badge..." or "once an exceptionally rare punishment"

- Four officers were decertified from 1999 to 2011. See bar chart below. 

> 2: "...dramatic increases" or "dramatic jump" or "the large increase in decertifications"

- Before the law change, the highest year was 17. After law change, each year has been 50 or more. See bar chart.

> 3: "Decertifications of Virginia law enforcement officers have jumped fourfold since lawmakers enacted reforms during the 2020 push for police accountability, according to a data analysis by the Virginia Center for Investigative Journalism at WHRO and MuckRock."

- "... jumped fourfold..." ---> 2021-2024 average $(50+81+94+81)/4 = 76.5$, 2020 = 17; $76/17 = 4.47$. See bar chart.

> 4: "Decertifications-or the revocation of a person's authority to serve as an officer-have averaged about 80 cases annually since the reforms were enacted in 2021."

- "... have averaged about 80 cases annually..." ---> $(50+81+94+81)/4 = 76.5$

> 6: "In the prior two decades, the highest number of officers decertified statewide per year was only 17..."

- See chart below.

> 7: "Between 1999 and 2011, just four officers in Virginia lost their certifications."

- See chart below

> In vis annotation: "From 1999 to 2009, only three officers were decertified."

- See chart below

```{r}
# Dataframe that sums the total of number of decertifications for each year since law change
by_year_present <- present %>% 
  mutate(year = year(fixed_date)) %>% 
  group_by(year) %>% 
  summarize(num = n()) %>% 
  # Remove two cases, one from 2025 since we don't have enough data and one from 2029 that is data error
  filter(!year %in% c("2025", "2029"))

# Calculate the average num of decertifications yearly since law change: 76.5
mean_present <- mean(by_year_present$num)


# Dataframe that sums the total of number of decertifications for each year before law change
by_year_prior <- prior %>% 
  mutate(year = year(fixed_date)) %>% 
  group_by(year) %>% 
  summarize(num = n()) 

# Dataframe of all years of certifications for comparison
merge_years <- rbind(by_year_present, by_year_prior) %>% 
  # Fill in zero for years not included in data: 2001, 2002, 2003, 2004, 2006, 2007, 2008, 2009, 2011
  complete(year = seq(min(year), max(year), by = 1),fill = list(num =0)) %>% 
  arrange(desc(year)) %>% 
  filter(year >= 2010)



# Export data to csv for Datawrapper chart
write.csv(merge_years, "../data/processed/vis/all_years.csv")


```
### Visualization in the article
<iframe title="2020 law created a big shift in Virginia's police decertifications" aria-label="Bar Chart" id="datawrapper-chart-xH2TV" src="https://datawrapper.dwcdn.net/xH2TV/4/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="765" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

> 9: "There are almost 25,000 certified law enforcement officers among the state's police departments, sheriff's offices and jails ... "

- Sum of all certified officers in table below = 24,980

```{r}
# Calculate the total number of certified officers
total_certified_officers <- sum(certified_directory$certified_officers_jailors, na.rm = TRUE)


```

## Data table of certified officers  
```{r}
#| output: true
datatable(certified_directory, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',
          rownames = FALSE, 

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))


```

# What agencies decertify the most officers? 
## Most agencies haven't decertified an officer since law change
### Sentences in the article 

> 10: "... most agencies haven't decertified a single officer since the law changed."

- "Most agencies..." ---> in data table below $263/373 = 0.705$ or about 70% and trend visualized in chart below (this chart will not be included in article)

> 11: "More than 260 agencies have not decertified any officers."

- 263 agencies that have a certified officer as of March 2025 have not decertified an officer 

> 12: "Of the agencies that have decertified officers, about half used the process just once."

- 56 agencies decertified one officer, 61 decertified more than one; $56/115 =  0.48$ or 48%


```{r}
# Dataframe of total number of decertifications at each agency that has decertified an officer since 2021
present_agency_totals <- present %>% 
  group_by(agency_name) %>% 
  summarize(decerts = n())


# 139 agencies don't have a certified officer, we'll exclude these later
no_certified_officer <- certified_directory %>% 
  filter(is.na(certified_officers_jailors))

# Dataframe of all agencies that have and haven't had a decertification since 2021, excluding those that don't have a certified officer as of March 2025
# Join to add decertifications column, NAs for zero decertifications 
certified_agencies <- certified_directory %>%
  filter(!is.na(certified_officers_jailors)) %>% 
  left_join(present_agency_totals, join_by(place_of_employment == agency_name)) 


# Dataframe of total number of decertifications at each agency that has decertified an officer before 2021
prior_agency_totals <- prior %>% 
  group_by(agency_name) %>% 
  summarize(num = n())

extra_agencies <- present_agency_totals %>% 
  anti_join(certified_agencies, join_by(agency_name == place_of_employment)) 

# Dataframe of agencies that haven't decertified an officer   
no_decertifications <- certified_agencies %>% 
  filter(is.na(decerts)) 


# 512 number of agencies in directory
# 139 number of agencies without a certified officer 
# 373 total agencies with at least one certified officer
# 115 agencies have decertified since 2021
# 263 no decertification 
# 5 agencies that did not have a certified officer as of March 2025, but did show up in our decertifications list
# ^ is the reason we have 263 with no decertification and not 258
  
```

## Histogram of agencies and decertifications 
```{r}
#| output: true
ggplot(certified_agencies, aes(x=decerts)) +
  geom_histogram(fill = "#fa8c00", boundary = 0) +
  labs(x = "Number of decertifications",
    y = "Number of agencies") +
  theme_classic()


  
```

```{r}

# Agencies with at least one certified officer 
at_least_one_certified_agencies <- certified_directory %>%
  filter(!is.na(certified_officers_jailors))

# Dataframe of number of decertifications grouped by how many agencies decertified that many officers
decert_bins <- present_agency_totals %>% 
  full_join(at_least_one_certified_agencies, join_by(agency_name == place_of_employment)) %>% 
  group_by(decerts) %>% 
  rename(num_decerts = decerts) %>% 
  summarize(num_agencies = n())

# Dataframe of agencies that decertified more than one officer 
greater_than_one_decert <- decert_bins %>% 
  filter(num_decerts > 1)

total_greater_than_one <- sum(greater_than_one_decert$num_decerts)
  
```

## Data table of number of agencies to number decertifications 
```{r}
#| output: true
datatable(decert_bins, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',
          rownames = FALSE, 

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))


```
## Some agencies have decertified a handful of officers or more
### Setences in the article 
> 8: "Five agencies have decertified more than ten officers: Virginia State Police, Chesapeake Police Department, Henrico County Sheriff's Office, Fairfax County Police Department and Fairfax County Sheriff's Office."

- Virginia State Police (16), Chesapeake (16), Henrico County SO (10), Fairfax County PD (10), Fairfax County SO (10). See data table below. 

> 13: "The Virginia State Police employs more officers than any other agency..."

-  Virginia State Police certified officers or jailors = 1,788. See data table below. 

> 14: "The Virginia State Police ... has decertified the highest number of officers - 16 - since the 2020 reforms were enacted."

- State police have decertified 16 officers, along with Chesapeake PD. See table below. 


> 15: "But the Chesapeake Police Department has 331 officers, just a fifth the size of the state police..."

- "... a fifth the size ... " ---> $331/1788 = 0.185$ or about 20%, a fifth. See data table below. 

> 16: "But the Chesapeake Police Department ... has also decertified as many officers - 16 - as the state police."

- See data table below. 

> 23: "Meanwhile, two of the state's largest agencies, Loudoun County Sheriff's Office and Virginia Beach Police Department ..."

- Loudon is the sixth largest force and Virginia Beach Police is the third. See data table below. 

> 24: "Loudoun County Sheriff's Office and Virginia Beach Police Department... have just two decertifications each."

- See data table below. 

```{r}
# Dataframe that combines number of decertifications with number of certified officers and compares rank and percentile rank, for agencies that have decertified an officer
present_agency_totals_cert <- present_agency_totals %>%
  left_join(certified_directory, join_by(agency_name == place_of_employment)) %>% 
  select(agency_name, num_decerts = decerts, certified_officers_jailors) %>% 
  mutate(percentile_decert = round(rank(num_decerts, ties.method = "min") / nrow(present_agency_totals) *100, digits = 2)) %>% 
  mutate(size_rank = rank(-certified_officers_jailors, ties.method = "min")) %>%
    mutate(percentile_size = round(rank(certified_officers_jailors, ties.method = "min") / nrow(present_agency_totals) *100, digits = 2)) %>% 
  mutate(decert_rank = rank(-num_decerts, ties.method = "min"))


# Dataframe that combines decertifications with certified officers and compares rank, for all 513 agences, not just those that have decertified 
certified_agencies_totals <- certified_agencies %>%
  rename(agency_name = place_of_employment) %>% 
  select(agency_name, num_decerts = decerts, certified_officers_jailors) %>% 
  mutate(size_rank = rank(-certified_officers_jailors, ties.method = "min")) %>%
  mutate(decert_rank = rank(-num_decerts, ties.method = "min"))


# Dataframe of top ten agencies have decertified and officer for Datawrapper chart
five_or_over <- present_agency_totals_cert %>% 
  filter(num_decerts >= 5)

top_ten <- present_agency_totals_cert %>% 
  slice_max(order_by = num_decerts, n = 10)
  
# Writing data to csv for Datawrapper
write.csv(five_or_over, "../data/processed/vis/agencies_five_or_over.csv")
write.csv(top_ten, "../data/processed/vis/top_ten.csv")
write.csv(present_agency_totals_cert, "../data/processed/vis/all_agencies_total_cert.csv")

```

## Data table of number of decertifications and number of certified officers in each agency
```{r}
#| output: true
datatable(certified_agencies_totals, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',
          rownames = FALSE, 

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))


```


> 17: "Chesapeake Police Department ... decertified only two officers in the two decades before the reforms."

- See data table below. 


> 18: "The Henrico County Sheriff's Office, which employs about 248 officers..."

- See data table above of number of decertifications and number of certified officers in each agency.

> 19: "The Henrico County Sheriff's Office ... as decertified 10 members of the force since the new law went into effect."

- See data table above of number of decertifications and number of certified officers in each agency.

> 20: "Before 2021, the Henrico County Sheriff's Office had not decertified a single office."

- See data table below; Henrico SO is not in table.

> 21: "Fairfax County Police Department, the second largest law enforcement agency in the state ..."

- Fairfax employs 1,270 officers, second only to state police. See table above. 

> 22: "Fairfax County Police Department ... has also decertified ten officers since the updated statute took effect."

- See table above. 

```{r}
# Dataframe of decertifications by agency prior to 2021
by_agency_prior <- prior %>% 
  group_by(agency_name) %>% 
  summarize(num = n()) 

```

## Data table of agencies that decertfied an officer before law changed in 2021
```{r}
#| output: true
datatable(by_agency_prior, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',
          rownames = FALSE, 

                            options = list(
                                paging = TRUE,
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))

```

### Visualization in the article 
<iframe title="Despite differences in size, some departments have decertified several officers since law changed" aria-label="Table" id="datawrapper-chart-JifBU" src="https://datawrapper.dwcdn.net/JifBU/3/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="400" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>



# What types of misconduct are most common?
## The big change since 2021 is lying and falsifying information
### Setences in the article 
> 25: "Since the reforms, the leading reasons for decertifications have been giving untruthful statements during an internal affairs investigation or falsifying documents."

- The causes are difficult to group because small variations, but even a rough grouping shows that the most common causes with 91 and 34 cases respectively are lying during an IA investigation and falsifying documents. See table below. 

> 26: "In Chesapeake, 14 of the 16 police officers lost their badges for lying..."

- See table below. 

> 27: "... all 10 deputies in the Henrico County Sheriff's Office were decertified for lying."

- See table below. 

```{r}
# Dataframe of grouped reasons for certifications 
type <- present %>% 
  group_by(reason) %>% 
  summarize(decerts = n()) 

# Decertifications at Chesapeake and Henrico as example 
type_henrico_chesapeake <-  present %>% 
  filter(agency_name %in% c("Chesapeake Police Department", "Henrico County Sheriff's Office")) 



```
## Data table of number of decertifications by decertification reason
```{r}
#| output: true
datatable(type, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',
          rownames = FALSE, 

                            options = list(
                                paging = TRUE,
                                pageLength = 10,
                                lengthMenu = c(5, 10, 25, 50),
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))
```
## Data table of officer level decertifications for Chesapeake PD and Henrico SO 
```{r}
#| output: true
datatable(type_henrico_chesapeake, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',
          rownames = FALSE, 

                            options = list(
                                paging = TRUE,
                                pageLength = 10,
                                lengthMenu = c(5, 10, 25, 50),
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))
```

# What about reinstatements?
## Reinstatements are also increasing
### Setences in the article 
> 5: "Reinstatements of decertified officers have also risen."

- Reinstatements have increased each year since 2021. See table below. 


> 28: "Reinstatements following appeals have also gradually increased since the new law took effect, reaching a high in 2024." 

- See table below. 

```{r}
# Dataframe of reinstatements by year
reinstated_year <- reinstated %>% 
  mutate(year = year(recert_date)) %>% 
  group_by(year) %>% 
  summarize(number = n())
```
## Data table of reinstatement totals by year
```{r}
#| output: true
datatable(reinstated_year, filter = 'top', class = 'cell-border stripe order-column', extensions = 'Buttons',
          rownames = FALSE, 

                            options = list(
                                paging = TRUE,
                                pageLength = 10,
                                lengthMenu = c(5, 10, 25, 50),
                                searching = TRUE,
                                fixedColumns = TRUE,
                                autoWidth = TRUE,
                                ordering = TRUE,
                                dom = 'tB',
                                buttons = c('copy', 'csv', 'excel')
                            ))
```